Why does preregistration increase the persuasiveness of evidence? A Bayesian rationalization.

### 2022-06-24 manuscript.Rmd
Update manuscript.Rmd

A successful prediction should lend more credibility to a theory than a postdiction, all else being equal.
The scientific community has long pondered the vital distinction between exploration and confirmation, discovery and justification, hypothesis-generating and hypothesis-testing and so forth [@hoyningen-hueneContextDiscoveryContext2006; @shmueliExplainPredict2010].
Confusing exploratory findings with confirmed theories has lead to a crisis of confidence in the results of empirical sciences [@ioannidisWhyMostPublished2005] and psychology in particular [@opensciencecollaborationEstimatingReproducibilityPsychological2015].
As a response, more and more researchers preregister their studies [@nosekPreregistrationRevolution2018] to honour this distinction and produce results that are considered confirmatory.

Indeed, rigorous application of preregistration prevents researchers from reporting a simple confirmatory story for a set of results produced by an arduous process of trial and error [@wagenmakersAgendaPurelyConfirmatory2012].
However, while many researchers intuitively recognize this as an advantage of preregistration, some remain unconvinced.
These skeptics question why a study should be preregistered at all.
Even researchers convinced of the advantages of preregistration face difficult problems when they exclusively rely on an intuitive appeal.
Specifically, an intuitive conviction is unable to address pragmatic questions, such as what to include in the preregistraion and how detailed this should be, or about the conditions that warrant deviating from it.
If the research community remains confused about the purpose of preregistration, it is bound to misapply it and hence decrease the efficiency of the scientific endeavour. 
We thus propose a principled justification for preregistration with a formal objective against which researchers can evaluate several pragmatic trade-offs.

@@

These issues arise from a fundamental confusion about the objective of preregistration.
Researchers are bound to be confused when they delegate an important scientific judgment to a simple decision rule, that only takes into account the existence of a preregistration.
Preregistering an inherently exploratory analysis (like testing dozens of relations) does not make it confirmatory, nor will a carefully conducted confirmatory study become exploratory if the researcher deviates from the preregistration in minor details.
Equating confirmatory research with preregistered research is only possible for studies that expect low type I error rate and will run/be analyzed without changes no matter what [@simmonsPreregistrationWhyHow2021; @bakkerEnsuringQualitySpecificity2020].
Under these conditions, the rule upholds, but such restricted use makes preregistration a niche solution unable to match the greater problem of replicability in psychological research or elsewhere.

We show that the simple decision rule is just a special case of a more general conceptualization under Bayesian reasoning. 
To that end, we first introduce some tools of Bayesian philosophy of science and map the exploration/confirmation distinction onto a dimensional quantity we call "theoretical risk" [a term borrowed from @meehlTheoreticalRisksTabular1978 but assigned to the probability of proving a hypothesis wrong, if it does not hold], which is inversely related to type I error rate.

We then outline two interpretations of how theoretical risk is impacted by preregistration.
The first interpretation corresponds to the traditional application of preregistration to research paradigms that focus on confirmation by maximizing the theoretical risk or equivalently by limiting type I error.
The second interpretation is our main contribution and demonstrates the broad applicability of preregistration for both exploratory and confirmatory studies that are implemented as preregistered or have undergone changes after preregistration.
Following this interpretation, the theoretical risk is not necessarily directly maximized by preregistration, but rather the uncertainty in judging the theoretical risk is minimized.

@@

Let us start by defining what we call expected epistemic value.
If researchers plan to conduct a study, they usually hope it will change their assessment of some theory's verisimilitude (truthlikeness).
In other words, they hope to learn something from conducting the study.
The amount of knowledge researchers gain from a particular study concerning the verisimilitude of a specific theory (which itself is an ontologocial concept) is what we call epistemic value.
Researchers cannot know what they will learn from a study before they have run it. However, they can develop an expectation that helps them to decide which study to conduct <!--AB in the sense of maximum information gain? What does "which study" mean? I guess you mean just any possible property of an study in the planning phase that can be altered?-->.
This expectation is what we term expected epistemic value.
To make our three arguments, we must assume three things about this estimation process and how it relates to choosing a study to conduct.

@@

The assumption of rationality can be connected to Bayesian reasoning and leads to our adoption of the framework.
Our rationale is as follows.
Researchers who decide to conduct a study are akin to choosing a study to bet on.
They have to "place the bet" by conducting the study, therefore, invest resources and stand to gain epistemic value with some probability.
This conceptualization of choosing a study as a betting problem allows us to apply a "Dutch Book" argument [@christensenCleverBookiesCoherent1991].
This argument states that any better must follow the axioms of probability to avoid being "irrational", i.e., accepting bets that lead to sure losses.
Fully developing a Dutch book argument for this problem requires careful consideration of what kind of studies to include as possible bets, defining a conversion rate from the stakes to the reward, and modelling what liberties researchers have in choosing a study.
Without deliberating these concepts further, we find it persuasive that researchers should not violate the axioms of probability if they have some expectation about what they stand to gain with some likelihood from conducting a study.
The axioms of probability are sufficient to derive the Bayes formula, on which we will heavily rely for our further arguments.
The argument is not sufficient, however, to warrant conceptualizing the kind of epistemic value we reason about in terms of probability; that remains a leap of faith.
Please note that our decision to adopt this aspect of the Bayesian philosophy of science does not imply anything about the statistical methods researchers use.
In fact, this conceptualization is intentionally reductionistic to be compatible with a wide range of philosophies of science and statistical methods researchers might subscribe to.

@@

Our first argument is that theorethical risk is crucial for judging theories.<!-- again: judging what?-->
Put simply, risky predictions create persuasive evidence if they turn out to be correct.
This point is crucial because we attribute much of the appeal of preregistration to this fact.

Let us make some simplifying assumptions and define notation.
We restrict ourselves to evidence of a binary nature (either exists or does not exist) <!--this seems like an ontological level but here again, aren't we on an epistemic level, so shouldn't it be: was observed or not? --> since continuous evidence would lead to some quite involved derivations.
<!--AB: What about measurement error? Is evidence thought to be measurement-free her? -->
We denote the probability <!--AB: personally, I think that a lot of confusion with Bayesian results is that the 'term' probability makes it sound objective and truthful; would you be okay with replacing probability with credibility to underline the subjective nature of this approach? Or would this be even more confusing?--> of a hypothesis before observing evidence as $P(H)$ and its complement as $P(¬H) = 1 - P(H)$.
<!--AB: Not all readers may understand what the probability of a hypothesis is. Elaborate more on the probability interpretation of Bayesians as credibility.-->
The probability of observing evidence under some hypothesis is $P(E|H)$.
We can calculate the probability of the hypothesis after observing the evidence with help from the Bayes formula:

@@

$P(H|E)$ is of great relevance since $P(H|E)$ is often used directly or indirectly as a measure of corroboration of a hypothesis.
In the tradition of Carnap (XXX), in its direct use, it is called corroboration as firmness; in its relation to the prior believe in the hypothesis, $P(H)$, it is called increase in firmness.
We refrain from discussing specific measures of corroboration since no measure shows universally better properties than others.
However, it is generally expected that any measure of corroboration increases monotonically with an increase in $P(H|E)$.

In short, to support our theory as good as possible, we want to increase $P(H|E)$. <!-- or: to corroborate or to confirm or...?-->
Increases in $P(H|E)$ are associated with increased epistemic value, of which we want to maximize the expectation. <!--Why the expectation? Because you take the expectation of an entire field or a scientist who does multiple experiments over her/his career? -->
So how can we increase $P(H|E)$?
The Bayes formula yields three components that influence corroboration, namely $P(H)$, $P(E|H)$ and $P(E)$.
The first option leads us to the unsurprising conclusion that a higher prior believe in a theory, $P(H)$, leads to higher $P(H|E)$.
If a hypothesis is more credible to begin with, the same increase in corroboration will lead to a higher credibility than if the study was less credible to begin with. <!--AB: not sure - state the obvious explicitly?-->
Anyhow, the a priori probability of a hypothesis is nothing our study design can change.
The second option is similar unsurprising; that is, an increase in $P(E|H)$ leads to higher $P(H|E)$.
In other words, researchers should ensure that their study design allows them to find evidence for their hypothesis, in case it is true.
While this truism is strongly related to study design, it is unrelated to preregistration.
Thus, $P(E)$ remains to be considered.
Since $P(E)$ is the denominator, increasing it will decrease $P(H|E)$: The more unlikely it is to observe evidence, the more it increases the probability of the hypothesis if we do observe it.
In other words, high risk, high reward.

@@

To better understand how preregistrations can do that, let us take a closer look at what contributes to $P(E)$.
Using the law of total probability, we can split $P(E)$ into two terms:

@@

We already have noted that there is not much to do about $P(H)$ (and hence its counter probability $P(¬H)$), and that it is common sense to increase $P(E|H)$.
<!--AB: from the perspective of a rational agent/researcher, of course, it makes sense to make P(E|H) large but isn't psychology facing a widespread problem of underpowered studies? I would make this very clear that this essentially relates to power (right?) and that historically we haven't been doing well on that account but that this problem is not pertinent to preregistration as such-->
The real lever to pull is, therefore, $P(E|¬H)$.
This probability tells us how likely it is that we find evidence in favor of the theory when in fact, the theory is not true.
Its counter probability $P(¬E|¬H)= 1 - P(E|¬H)$ is what we call "theoretical risk", because it is the risk a theory takes on in predicting the occurrence of particular evidence in its favour.
We "borrow" the term from @meehlTheoreticalRisksTabular1978, though he has not assigned it to the probability $P(¬E|¬H)$.
In a response to the original paper, @kuklaClinicalStatisticalTheory1990 argued that the arguments layed out in @meehlTheoreticalRisksTabular1978 can be reconstructed in purely Bayesian framework.
However, while he did not name $P(¬E|¬H)$ but did suggest that @meehlTheoreticalRisksTabular1978 used the term "very strange coincidence" for a small $P(E|¬H)$ which would imply that $P(¬E|¬H)$ can be related to or even equated to theoretical risk.

@@

<!--AB can we add an example that explains this whole idea once more but with a concrete example? I think that the probability of evidence given NOT H is just difficult to wrap one's head around. It would be ideal to illustrate this with an example that presents both a weak and a strong P(E given not H).

I have a hard time coming up with an easy example that relates this simplified case
to a problem of psychological research.

H: a depression treatment is effective
not H: the treatment is not effective

In a weak risk case, the prediction is fuzzy, P(E | not H) is large, theoretical risk is small:

E: people have some depression score (other than zero) after the treatment
not E: people have exactly zero depression scores

In a strong risk case, the prediction is precise, P(E | not H) is small, theoretical risk is large
E: people improve at least 10 points on the depression scale
not E: people improve less or equal to 10 points on the depression scale
-->

Let us note some interesting properties of theoretical risk, $P(¬E|¬H)$.
<!--AB: I rephrased this sentence. Is credibility equal to epistemic value?-->
First, if the objective is to increase the credibility of a theory $P(H|E)$, this can
be achieved by increasing theoretical risk.
Second, if the theoretical risk is smaller than $P(E|H)$ it follows that $P(H|E)$ must decrease when observing the evidence.
<!--AB: explain more what this means. When is this condition true in reality?-->
Third, if the theoretical risk equals zero, then $P(H|E)$ can at best be equal to $P(H)$ but only if $P(H|E)$ is one.
In other words, observing a sure fact does not lend credence to a hypothesis.

<!--Is this the word "this" refering to the third point only or all three?-->
This sounds like a truism but is directly related to Popper's seminal criterion of demarcation.
He stated that if it is impossible to prove a hypothesis false ($P(E|¬H) = 1$ or $P(¬E|¬H) = 0$), it can not be considered a scientific hypothesis [@popperLogicScientificDiscovery2002, p. 18].
We note these relations to underline that the Bayesian rational we apply here is able to reconstruct many commonly held views on riskiness and epistemic value.

@@

After we discussed that increasing the theoretical risk will increase the epistemic value, it is intuitive to task preregistration with maximizing theoretical risk.
Indeed, limiting the false positive (or, type I error rate) is commonly stated as a goal of preregistration. <!--AB: citation needed; maybe verbose quote?-->
We argue that while such a conclusion is plausible, we must first consider at least two constraints that place an upper bound on the theoretical risk.

<!--AB: the kind of more concrete example that you give here is needed much earlier when you define the different conditional probabilities; maybe even in the very beginning, it could be helpful to give examples of risky vs non-risky predictions in psychology from published literature?!-->
First, the theory itself limits theoretical risk:
Some theories simply do not make risky predictions, and preregistration will not change that.
Consider the case of a researcher contemplating the relation between two sets of variables.
Suppose each set by itself is well studied, and strong theories tell the researcher how the variables within each of the sets relate to each other.
Now, our imaginary researcher considers the relation between these two sets.
For lack of a better theory, they assume that some relation between any variables of the two sets exists.
This is not a risky prediction to make in psychology, even without statistical issues like alpha inflation due to multiple testing [@orbenCrudReDefined2020].
However, we would consider it a success if the researcher would use the evidence to develop a more precise (and therefore risky) theory, e.g. by specifying which variables from one set relate to which variables from the other set, to what extent, in which direction, with which functional shape etc.
We will later show that preregistration increases the belief one can stake in the further specified theory, though it remains low until substantiated by testing it again. <!--AB: "till" is never used in formal writing-->
The point, however, is that we want to show that preregistration increases the expected epistemic value independent of the theory being tested.

Second, available resources limit theoretical risk.
Increasing $P(¬E|¬H)$ will usually decrease $P(E|H)$ unless more resources are invested. <!-- I recommend giving specific names to each of the conditional probabilities and then refer to those names and have the math notation in parentheses; then, this becomes easier to read and understand, e.g.,

Increasing theoretical risk ($P(¬E|¬H)$) will usually decrease the probability of observing the evidence under the hypothesis ($P(E|H)$).

Is there a term, we can introduce for $P(E|H)$, like power?

-->
In other words, one can not increase power while maintaining the same type I error rate without increasing the invested resources.
Tasking preregistration with an increase in theoretical risk makes it difficult to balance this trade-off.
Mindlessly maximizing theoretical risk would either never produce evidence or require huge amounts of resources.

@@

Preregistration is often discussed in relation to statistical methods, so let us note some relations between $P(E|¬H)$ and statistical methods.
If you consider the overly simplistic case where the research hypothesis is equal to the statistical alternative hypothesis, then the nill-hypothesis is $¬H$, and therefore $P(E|¬H)$ is equivalent to the type I error rate.
Researchers who choose a smaller type I error rate can hence be more sure of their results, if significant.
In other words, the probability that a given significant finding reflects a true effect is larger if researchers lower their type I error rates.
However, the research hypothesis is seldomly equal to the statistical null hypothesis.
We argue that $P(E|¬H)$ (and hence its inverse, theorethical risk) also encompasses factors outside the statistical realm, most notably the study design and broader analytical strategies. <!--AB: can you elaborate more?!-->

@@

## Sources of Uncertainty <!--AB really causes?-->

As we noted, it is possible to quantify how statistical models affect the theoretical risk based on mathematical considerations and simulation.
However, other factors in the broader context of a given study are much harder to quantify.
If one chooses to focus only on the contribution of statistical methods to $P(E|¬H)$, one is bound to underestimate it.
Take, for example, a t-test of mean differences in two samples.
Under ideal circumstances (assumption of independence, normality of residuals, equal variance), it will keep its promised type I error rate.
However, researchers might do many very reasonable things in the broader context of the study that affect theoretical risk:
They might exclude outliers, choose to drop an item before computing a sum score, broaden their definition of the population to be sampled (thus jeopardizing distributional assumptions<!-- AB: is this what you meant?-->), translate their questionnaires into a different language without validating it, impute missing values, or any number of other things.
All of these decisions carry a small risk that they increase the likelihood of obtaining evidence despite the underlying research hypothesis being false.
Even if the t-test itself perfectly maintains its type I error rate, these factors influence $P(E|¬H)$.
While, in theory, these factors may leave $P(E|¬H)$ unaffected or even decrease it, we argue that this is not the case in practice.
Whether researchers want to or not, except under strict blinding, they continuously process information about how the study is going.
While one can hope that processing this information does not affect their decision making either way, this cannot be ruled out.
We, therefore, conclude that statistical properties only guarantee a lower bound to the quantity we seek to minimize.
The only thing we can conclude with some certainty is that $P(E|¬H)$ is not lower than what the statistical model guarantees without knowledge about other factors that play a role.

@@

Neither absolute certainty nor uncertainty are realistic scenarios but represent the boundary conditions that all realistic conditions fall in.
Depending on the distribution that expresses our state of knowledge about the theoretical risk a study took on, our expectation for what to gain varies considerably.
Uncertainty about theoretical risk is expressed through the entropy of the distribution. Under Gaussian assumptions, this is proportional to the variance of the distribution.
Generally, we expect that increases in uncertainty (expressed as more entropic distributions) lead to a decreased expected epistemic value.

<!--AB: I am confused that a new section heading is following but I haven't still understood or seen what follows from the expectation that you just gave. I assume the problem is that you created Figures 1 & 2 to illustrate the consequences of the choices you made and the expectation you presented but you do not refer to these Figures from the text, o the reader is lost and may never read the figures... please refer to the figures and explain what is going on in the text. Currently, I do not know what to take from Figure 2 other than that there are different ways to formalize this. Please also cite all sources for these measures in the text.-->

@@

We hope to have persuaded the reader to accept two arguments: First, the theoretical risk is important for judging evidential support for theories.
Second, the theoretical risk is inherently uncertain and the degree of uncertainty diminishes the persuasiveness of the gathered evidence. <!--AB my rewrite correct?-->
The last argument is that preregistrations reduces this uncertainty.

@@

Communicating clearly what the authors did to arrive at the evidence (<!--AB: "to arrive at the evidence" is unclear to me, do you mean a conditional probability or really the evidence itself, say, in numeric form in a data table)? --> is crucial for judging the theoretical risk they took.
Preregistrations are ideal to communicate just that, because any description after the fact is suspect to be incomplete.
The authors could have decided to exclude a number of analytic strategies they tried out etc.
That is not to say that any study that was not preregistered was subjected to practices of p hacking.
The point is, we can not exclude this and a myriad of other possibilities and, hence, are left uncertain.
If the researcher do describe what they intend to do beforehand, and then report they did exactly that, the readers can be certain, that they have received a complete account of the situation.
They still might be uncertain about the actual theoretical risk the authors took but much less so.
Remaining sources of uncertainty might be unfamiliarity with statistical methods or experimental paradigms, the probability of plain error in analyses etc.
In any case a well written preregistration should aim to reduce the uncertainty about the theoretical risk and hence increase the persuasiveness of evidence.

>>>>

A successful prediction should lend more credibility to a theory than a postdiction, all else being equal.
The scientific community has long pondered the vital distinction between exploration and confirmation, discovery and justification, hypothesis-generating and hypothesis-testing and so forth [see @hoyningen-hueneContextDiscoveryContext2006].
Confusing exploratory findings with confirmed theories has lead to a crisis of confidence in the results of empirical sciences [@ioannidisWhyMostPublished2005] and psychology in particular [@opensciencecollaborationEstimatingReproducibilityPsychological2015].
As a response, more and more researchers preregister their studies [@nosekPreregistrationRevolution2018] to honour this distinction and produce results that are considered confirmatory.
Indeed, rigorous application of preregistration prevents researchers from reporting a simple confirmatory story for a set of results produced by an arduous process of trial and error [@wagenmakersAgendaPurelyConfirmatory2012].
However, while many researchers intuitively recognize this as an advantage of preregistration, some remain unconvinced.
These skeptics question why a study should be preregistered at all.
Even researchers convinced of the advantages of preregistration face difficult problems when they exclusively rely on an intuitive appeal.
Specifically, an intuitive conviction is unable to address pragmatic questions, such as what to include in the preregistraion and how detailed this should be, or about the conditions that warrant deviating from it.
If the research community remains confused about the purpose of preregistration, it is bound to misapply it and hence decrease the efficiency of the scientific endeavour. 
We thus propose a principled justification for preregistration with a formal objective against which researchers can evaluate several pragmatic trade-offs.

@@

These issues arise from a fundamental confusion about the objective of preregistration.
Researchers are bound to be confused when they delegate an important scientific judgment to a simple decision rule, that only takes into account the existence of a preregistration.
Preregistering an inherently exploratory analysis (like testing dozens of relations) does not make it confirmatory, nor will a carefully conducted confirmatory study become exploratory if the researcher deviates from the preregistration in minor details.
Equating confirmatory research with preregistered research is only possible for studies that expect low type I error rate and will run/be analyzed without changes no matter what [@simmonsPreregistrationWhyHow2021; @bakkerEnsuringQualitySpecificity2020].
Under these conditions, the rule upholds, but such restricted use makes preregistration a niche solution unable to match the greater problem of replicability in psychology and elsewhere.

We show that the simple decision rule is just a special case of a more general conceptualization under Bayesian reasoning. 
To that end, we first introduce some tools of Bayesian philosophy of science and map the exploration/confirmation distinction onto a dimensional quantity we call "theoretical risk" [we borrow the term from @meehlTheoreticalRisksTabular1978], which is inversely related to type I error rate.
We then outline two interpretations of how theoretical risk is impacted by preregistration.
The first interpretation corresponds to the traditional application of preregistration to research paradigms that focus on confirmation by maximizing the theoretical risk or equivalently by limiting type I error.
The second interpretation is our main contribution and demonstrates the broad applicability of preregistration for both exploratory and confirmatory studies that are implemented as preregistered or have undergone changes after preregistration.
Following this interpretation, the theoretical risk is not necessarily directly maximized by preregistration, but rather the uncertainty in judging the theoretical risk is minimized.

@@

Let us start by defining what we call expected epistemic value.
If researchers plan to conduct a study, they usually hope it will change their assessment of some theory's verisimilitude (truthlikeness).
In other words, they hope to learn something from conducting the study.
The amount of knowledge researchers gain from a particular study concerning the verisimilitude of a specific theory is what we call epistemic value.
While researchers can not know what exactly they will learn from a study, they can form an expectation that helps them decide which study to conduct.
This expectation is what we term expected epistemic value.
To make our three arguments, we must assume three things about this estimation process and how it relates to choosing a study to conduct.

@@

The assumption of rationality can be connected to Bayesian reasoning and leads to our adoption of the framework.
Our rationale is as follows.
Researchers who decide to conduct a study are akin to choosing a study to bet on.
They have to "place the bet" by conducting the study, therefore, invest resources and stand to gain epistemic value with some probability.
This conceptualization of choosing a study as a betting problem allows us to apply a "Dutch Book" argument [@christensenCleverBookiesCoherent1991].
This argument states that any better must follow the axioms of probability to avoid being "irrational", i.e., accepting bets that lead to sure losses.
Fully developing a Dutch book argument for this problem requires careful consideration of what kind of studies to include as possible bets, defining a conversion rate from the stakes to the reward, and modelling what liberties researchers have in choosing a study.
Without deliberating these concepts further, we find it persuasive that researchers should not violate the axioms of probability if they have some expectation about what they stand to gain with some likelihood from conducting a study.
The axioms of probability are sufficient to derive the Bayes formula, on which we will heavily rely for our further arguments.
The argument is not sufficient, however, to warrant conceptualizing the kind of epistemic value we reason about in terms of probability; that remains a leap of faith.
Please note that our decision to adopt this aspect of the Bayesian philosophy of science does not imply anything about the statistical methods researchers use.
In fact, this conceptualization is purposefully reductionistic to be compatible with a wide range of philosophies of science and statistical methods researchers might subscribe to.

@@

Our first argument is that theorethical risk is crucial for judging theories.
Put simply, risky predictions create persuasive evidence if they turn out to be correct.
This point is crucial because we attribute much of the appeal of preregistration to this fact.

Let us make some simplifying assumptions and define notation.
We restrict ourselves to evidence of a binary nature (either exists or does not) since continuous evidence would lead to some quite involved derivations.
We denote the probability of a hypothesis before observing evidence as $P(H)$ and its complement as $P(¬H) = 1 - P(H)$.
The probability of observing evidence under some hypothesis is $P(E|H)$.
We can calculate the probability of the hypothesis after observing the evidence with help from the Bayes formula:

@@

The posterior probability ($P(H|E)$) is of great relevance since it is often used directly or indirectly as a measure of corroboration of a hypothesis.
In the tradition of Carnap (XXX), in its direct use, it is called corroboration as firmness; in its relation to the a priori probability ($P(H)$), it is called increase in firmness.
We refrain from discussing specific measures of corroboration since no measure shows universally better properties than others.
However, it is generally expected that any measure of corroboration (and even increases monotonically with an increase in posterior probability ($P(H|E)$).

In short, we want to increase posterior probability ($P(H|E)$).
Increases in posterior probability ($P(H|E)$) are associated with increased epistemic value, of which we want to maximize the expectation.
So how can we increase posterior probability?
The Bayes formula gives us three options to investigate, namely $P(H)$, $P(E|H)$ and $P(E)$.
The first option leads us to the unsurprising conclusion that higher a priori probability ($P(H)$) leads to higher posterior probability ($P(H|E)$).
However, the prior probability of a hypothesis is nothing our study design can change.
The second option is similar commonsensical; that is, an increase in $P(E|H)$ leads to higher posterior probability ($P(H|E)$).
$P(E|H)$ is the probability of obtaining evidence for a theory, when the theory holds.
In other words, how probable is it that we "detect" that a theory holds, we therefore call it "detectability".
Consequently, researchers should ensure that their study design allows them to find evidence for their hypothesis, in case it is true.
When applied strictly within the bounds of null hypothesis testing, detectability is equivalent to power (or the inverse of type II error rate).
However, while detectability is of great importance for study design, it is not directly relevant for the objective of preregistration.
Thus, $P(E)$ remains to be considered.
Since $P(E)$ is the denominator, increasing it will decrease the posterior probability: The more unlikely it is to observe evidence, the more it increases the probability of the hypothesis if we do observe it.
In other words, high risk, high reward.

@@

To better understand how preregistrations can achieve that, let us take a closer look at what factors contribute to $P(E)$. <!--AB: is this expectedness (Pap, 1962) as quoted on p.818 in Meehl's risk paper?-->
Using the law of total probability, we can split $P(E)$ into two terms:

@@

We already have noted that there is not much to do about prior probability ($P(H)$, and hence its counter probability $P(¬H)$), and that it is common sense to increase detectability ($P(E|H)$).
The real lever to pull is, therefore, $P(E|¬H)$.
This probability tells us how likely it is that we find evidence when in fact, the theory is not true.
Its counter probability $P(¬E|¬H)= 1 - P(E|¬H)$ is what we call "theoretical risk", because it is the risk a theory takes on in predicting the occurrence of particular evidence in its favour.

Let us note some interesting properties of theoretical risk ($P(¬E|¬H)$).
First, increasing theoretical risk leads to higher posterior probability ($P(H|E)$, our objective).
Second, if the theoretical risk is smaller than detectability ($P(E|H)$) it follows that the posterior probability must decrease when observing the evidence.
If detectability exceeds theorethical risk, the evidence is less likely under the theory than it is when the theory does not hold.
Third, if the theoretical risk equals zero, then posterior probability is at best equal to prior probability but only if detectability is perfect ($P(H|E)$ = 1).
In other words, observing a sure fact does not lend credence to a hypothesis.

This sounds like a truism but is directly related to Popper's seminal criterion of demarcation.
He stated that if it is impossible to prove a hypothesis false ($P(¬E|¬H) = 0$, theoretical risk is zero), it can not be considered a scientific hypothesis [@popperLogicScientificDiscovery2002, p. 18].
We note these relations to underline that the Bayesian rational we apply here is able to reconstruct many commonly held views on riskiness and epistemic value.

Both theorethical risk ($P(¬E|¬H)$) and detectability ($P(E|H)$) aggregate uncountable influences, otherwise they could not model the process of evidential support for theories.
To illustrate the concepts we introduced up to here, consider the following  example of a single theory and three experiments that may test it. 
The experiments were created to illustrate how they may differ in their theoretical risk and detectability.
Suppose the primary theory is about the cognitive phenomenon of "insight".
For the purpose of illustration, we define it somewhat hand-wavinlyg as an cognitive abstraction that allows agents to consistently solve a well-defined class of problems.
We pose the hypothesis that the following problem belongs to such class of insight problems:

> Use five matches (Ⅰ Ⅰ Ⅰ Ⅰ Ⅰ) to form the number eight.

We propose three experiments that differ in theoretical risk and detectability.
All experiments take a sample of ten psychology students.
We present the students the problem for a brief span of time.
After that, the three experiments differ as follows:

1. the experimenter gives a hint that  the problem is easy to solve when using Roman numerals; if all students come up with the solution, she records it as evidence for the hypothesis.
2. the experimenter shows the solution "Ⅷ" and explains it; if all students come up with the solution, she records it as evidence for the hypothesis.
3. the experimenter does nothing; if all students come up with the solution, she records it as evidence for the hypothesis.

We argue that experiment 1 has high theoretical risk ($P(¬E_1|¬H)$) and high detectability ($P(E_1|H)$).
If "insight" has nothing to do with solving the problem ($¬H$), than presenting the insight that roman literals might be used, should not lead to all students solving the problem ($¬E_1$); the experiment has therefore high theoretical risk ($P(¬E_1|¬H)$)
Conversly, if insight is required to solve the problem ($H$), than it is probable to help all students to solve the problem ($E_1$); the experiment has therefore high detectability ($P(E_1|H)$).
The second experiment, on the other hand, has low theoretical risk ($P(¬E_2|¬H)$).
Even if "insight" has nothing to do with solving the problem ($¬H$), there are other plausible reasons for observing the evidence ($E_2$), because the students could simply copy the solution, without having any insight.
With regard to detectability experiment 1 and 2 differ in no obvious way.
Experiment 3, however, also has low detectability.
It is unlikely that all students come up with the correct solution in a short time ($E_3$), even if insight is required ($H$); the experiment 3 has therefore low detectability ($P(E_3|H)$).
The theoretical risk, however, is also low in absolute terms but high compared to the theoretical risk.
In the unlikely event that all 10 students lay the matches down in the form of the roman numeral Ⅷ ($E_3$) it is probably due to insight ($H$) and not by chance $P(¬E_2|¬H)$).
Of course, in practice, we would allow the evidence to be probabilistic, e.g., relax the requirement of "all students" to nine out of ten students, more than eight, and so forth.
As argued earlier, the remainder of the paper will focus on binary, non-probabilisitic evidence to keep the mathematical notation as simple as possible.
We discuss the relation between statistical methods and theoretical risk in the section [Statistical Methods].

@@

@@

After we discussed that increasing the theoretical risk will increase the epistemic value, it is intuitive to task preregistration with maximizing theoretical risk.
Indeed, limiting type I error rate is commonly stated as a goal of preregistration.
We argue that while such a conclusion is plausible, we must first consider at least two constraints that place an upper bound on the theoretical risk.

First, the theory itself limits theoretical risk:
Some theories simply do not make risky predictions, and preregistration will not change that.
Consider the case of a researcher contemplating the relation between two sets of variables.
Suppose each set by itself is well studied, and strong theories tell the researcher how the variables within the set relate.
However, our imaginary researcher considers the relation between these two sets.
For lack of a better theory, they assume that some relation between any variables of the two sets exists.
This is not a risky prediction to make in psychology, even without statistical issues like alpha inflation [@orbenCrudReDefined2020].
However, we would consider it a success if the researcher would use the evidence to develop a more precise (and therefore risky) theory, e.g. by specifying which variables from one set relate to which variables from the other set, to what extent, in which direction, with which functional shape etc.
We will later show that preregistration increases the belief one can stake in the further specified theory, though it remains low till substantiated by testing it again.
The point, however, is that we want to show that preregistration increases the expected epistemic value without regard to the theory being tested.

Second, available resources limit theoretical risk.
Increasing theorethical risk ($P(¬E|¬H)$) will usually decrease detectability ($P(E|H)$) unless more resources are invested.
In other words, one can not increase power while maintaining the same type I error rate without increasing the invested resources.
Tasking preregistration with an increase in theoretical risk makes it difficult to balance this trade-off.
Mindlessly maximizing theoretical risk would either never produce evidence or require huge amounts of resources.

@@

Preregistration is often discussed in relation to statistical methods, so let us note some relations between $P(E|¬H)$ (the inverse of theorethical risk) and statistical methods.
If you consider the overly simplistic case where the research hypothesis is equal to the statistical alternative hypothesis, then the nill-hypothesis is $¬H$, and therefore $P(E|¬H)$ is equivalent to the type I error rate.
Researchers who choose a smaller type I error rate can hence be more sure of their results, if significant.
However, the research hypothesis is seldomly equal to the statistical null hypothesis.
We argue that $P(E|¬H)$ (and hence theorethical risk) also encompasses factors outside the statistical realm, most notably the study design and broader analytical strategies.

@@

## Causes of Uncertainty

As we noted, it is possible to quantify how statistical models affect the theoretical risk based on mathematical considerations and simulation.
However, other factors in the broader context of the study are much harder to quantify.
If one chooses to focus only on the contribution of statistical methods to $P(E|¬H)$, one is bound to underestimate it.
Take, for example, a t-test.
Under ideal circumstances (assumption of independence, normality of residuals, equal variance), it stays true to its type I error rate.
However, researchers might do many very reasonable things in the broader context of the study that affect theoretical risk:
They might exclude outliers, choose to drop an item, enlarge their definition of the population to be sampled, translate their questionnaires, impute missing values, or any number of other things.
All of these decisions carry a small risk that they increase the likelhood of obtaining evidence despite the underlying hypothesis being false.
Even if the t-test itself perfectly maintains its type I error rate, these factors must be added to $P(E|¬H)$.
While, in theory, these factors may leave $P(E|¬H)$ unaffected or even decrease it, we argue that this is not the case in practice.
Whether researchers want to or not, except under strict blinding, they continuously process information about how the study is going.
While one can hope that processing this information does not affect their decision making either way, this cannot be secured.
We, therefore, conclude that statistical properties only guarantee a lower bound to the quantity we seek to minimize.
The only thing we can conclude with some certainty is that $P(E|¬H)$ is not lower than what the statistical model guarantees without knowledge about the other factors at play.

@@

Neither absolute certainty nor uncertainty are realistic scenarios but represent the boundary conditions that all realistic conditions fall in.
Depending on the distribution that expresses our state of knowledge about the theoretical risk a study took on, our expectation for what to gain varies considerably.
Uncertainty about theoretical risk is expressed through the variance (or rather entropy) of the distribution.
Generally, we expect that increases in uncertainty (expressed as more entropic distributions) lead to a decreased expected epistemic value.

@@

We hope to have persuaded you to accept two arguments: First, the theoretical risk is important for judging evidential support for theories.
Second, the theoretical risk is inherently uncertain, which diminishes the persuasiveness of the gathered evidence.
The last argument is that preregistrations reduces this uncertainty.

@@

Communicating clearly what the authors did to arrive at the evidence is crucial for judging the theoretical risk they took.
Preregistrations are ideal to communicate just that, because any description after the fact is suspect to be incomplete.
The authors could have decided to exclude a number of analytic strategies they tried out etc.
That is not to say that any study that was not preregistered was subjected to practices of p hacking.
The point is, we can not exclude this and a myriad of other possibilities and, hence, are left uncertain.
If the researcher do describe what they intend to do beforehand, and then report they did exactly that, the readers can be certain, that they have received a complete account of the situation.
They still might be uncertain about the actual theoretical risk the authors took but much less so.
Remaining sources of uncertainty might be unfamiliarity with statistical methods or experimental paradigms, the probability of plain error in analyses etc.
In any case a well written preregistration should aim to reduce the uncertainty about the theoretical risk and hence increase the persuasiveness of evidence.
